### 2020-07-07 Model Data Sheets and Code/Zaca MCF Mortality Model Results-DESKTOP-KM0MQ9O.Rmd
more model trials, updates/new data, prelim results

mcf <- read.csv('Zaca MCF variables for model.csv') %>% 
  mutate(mortality_qual= fct_relevel(mortality_qual, levels=c("None", "Low", "Moderate", "High"))) %>% 
  # mutate(dnbr_qualitative = case_when(
  #   dnbr < 0.1 ~ "unburned",
  #   dnbr < 0.27 ~ "low",
  #   dnbr < 0.66 ~ "moderate",
  #   dnbr > 0.66 ~ "high")) %>% 
  # mutate(dnbr_qualitative = as.factor(dnbr_qualitative),
  #        dnbr_qualitative = fct_relevel(dnbr_qualitative, levels=c("high", "moderate", "low", "unburned"))) %>% 
  mutate(aspect_transformed = (1-cos((2*pi*aspect)/360))/2) %>% 
  mutate(three_class_mortality = case_when(
    mortality_qual %in% "High" ~ "High",
    mortality_qual %in% "Moderate" ~ "Moderate",
    mortality_qual %in% c("Low", "None") ~ "Low")) %>% 
  mutate(three_class_mortality = as.factor(three_class_mortality)) %>% 
  mutate(three_class_mortality= fct_relevel(three_class_mortality, 
                                     levels=c("Low", "Moderate", "High"))) %>% 
  mutate(tc_mortality = as.factor(tc_mortality))

@@

@@

### Methods background

using forced sample sizes (see https://stats.stackexchange.com/questions/457451/how-r-randomforest-sampsize-works) because the distribution of observations in each mortality class is very uneven and results in a model that severely underpredicts observations in the mortality class with few observations. Bc of the small sample size, we will not be using training and test data (and bc RF basically creates its own train/test data).

Different RF trials:

  - Mixed confier forests
    + response variable: type conversion
    + response variable: 4 mortality classes (none, low, momderate, high)
    + response variable: 3 mortality classes (none/low, moderate, high)
    + response variable: 2 mortality classes (none/low/moderate, high)
    + response variable: rdnbr
  - Coulter pines 
    + response variable: rdnbr
  - mixed conifer forests and coulter pines combined
    + response variable: rdnbr

Note: manual classification of mortality and type conversion for the coulter pines was not done. Aint no one got time for that

The type conversion mortality classes-
tc_mortality 0 = "not_tc" (n=99)
tc_mortality 1 = type converted (n=>600)

### Goals/Questions

1. Can the mcf model distinguish between type conversion and the other mortality types? 
  - so far, no! so variable importance is not meaningful for that model

2. Which variables contribute to type conversion of MCFs?

3. How well does rdnbr compare to manual estimates of mortality?

4. How does using rdnbr and manually derived mortality compare as a response variable?

5. Which variables contribute to high/low mortality? high/low rdnbr?

5. Does conifer type contribute to different variable importances?

@@

ggplot(mcf, aes(x=ndvi, y=rdnbr_2007)) +
  geom_point()

ndvi_rdnbr_corr <- lm(rdnbr_2007~ndvi, data=mcf)
summary(ndvi_rdnbr_corr)

@@

### RF manual mortality: 4 classes (none, low, moderate, high)

Just using for loop to generate several graphs so its easier to compare them and make sure importance is stable
```{r}

for (i in 1:10) {

rfm <- randomForest(tc_mortality ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + vpd_jan + drainage_association + ndvi, 
                    data = mcf, 
                    sampsize = c('1' = 99, '0' = 99), 
                    ntree=1500, 
                    importance=T,
                    replace = T)
# OOB and importance
rfm
importance(rfm)
#plot to visualize variable importance
varImpPlot(rfm)

```
variable importance seems stable AND class accuracy is high and equal between classes!!

#### ROC and AUC
```{r}
# code: https://stats.stackexchange.com/questions/188616/how-can-we-calculate-roc-auc-for-classification-algorithm-such-as-random-forest
rfm_roc <-  roc(mcf$tc_mortality, rfm$votes[,1])
plot(rfm_roc) # plot comes out a little weird
auc(rfm_roc)

## PDP v. ALE

```{r}
#PDP
a <- partialPlot(rfm, as.data.frame(mcf), ndvi, "0") #0=not type converted
b <- partialPlot(rfm, as.data.frame(mcf), ndvi, "1", col="red") #1=type converted
lines(a, col="green")

a <- partialPlot(rfm, as.data.frame(mcf), slope, "0") #0=not type converted
b <- partialPlot(rfm, as.data.frame(mcf), slope, "1", col="red") #1=type converted
lines(a, col="green")

a <- partialPlot(rfm, as.data.frame(mcf), sr, "0") #0=not type converted
b <- partialPlot(rfm, as.data.frame(mcf), sr, "1", col="red") #1=type converted
lines(a, col="green")

#ale (see detailed notes in ALE Plots code chunk below; not working right now)
train_x <- mcf[-which(names(mcf) == "tc_mortality")]
ndvi_predictor_notTC <- Predictor$new(rfm, data = train_x, y = mcf$tc_mortality, type = "prob", class = "0")
ndvi_not <- plot(FeatureEffect$new(ndvi_predictor_notTC, feature = "ndvi")) + ggtitle("not TC")

#predict values for the other mortality classes for the same variable
ndvi_predictor_tc <- Predictor$new(rfm, data = train_x, type = "prob", class = "1")
sr_moderate <- plot(FeatureEffect$new(ndvi_predictor_tc, feature = "ndvi")) + ggtitle("TC")
```
so looking at the 3 most important variables for type conversino plots, all of the probabilities are below zero. only the not type converted stands have probabilities above 0, with some values of the explanatory variable having very high probabilities. So what this tells me is the model CANNOT distinguish type converted stands from non type converted stands.

### Random Forest and forced sample sizes with training and test data

hmmmm. not sure how to do this with training and test data since the train/test data frames will have different quantities of the type conversion classes each time. if i do it the way I have been doing it below, then variable importance is too unstable. even using my equal distribution sample size code and adding in the sampsize function results in unstable variable importance. 
```{r}
#train and test data 
sample <- sample(1:nrow(mcf), size = 0.75*nrow(mcf))
train <- mcf[sample, ]
test  <- mcf[-sample, ]

for (i in 1:10) {
rfm <- randomForest(tc_mortality ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + vpd_jan + drainage_association + ndvi, 
                    data = mcf, 
                    sampsize = c('1' = 99, '0' = 99), 
                    ntree=1500, 
                    importance=T,
                    replace = T)
rfm

#plot to visualize variable importance
varImpPlot(rfm)
#model_importance <- importance(rfm)
importance(rfm)
}
#predcit test data
pp<-predict(rfm, test)

#test data accuracy
tabb<- table(pp, test$tc_mortality)
acc<-1-sum(diag(tabb))/sum(tabb) # % missclassified 
test_accuracy <-(1-acc)*100 #model accuracy
```

@@

#random forest code---------------------------------------------------
for (i in 1:15)
{

@@

  #random forest model
  rfm <- randomForest(tc_mortality ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + ridge + drainage_association + ndvi, data = train, ntree=1500, importance=T, nscale=0, sampsize = c('tc' = 75, 'not_tc' = 75), replace = T)

@@

  # lists to save model outputs
  # test_accuracy_list <- append(test_accuracy_list, test_accuracy)
  # write.table(model_importance, file="rfm model tc importance.csv", sep=",",col.names=NA, append=T) 
  # write.table(tabb, file = "mcf tc test accuracy table.csv", sep=",",col.names=NA, append=T)
}

>>>>

mcf <- read.csv('Zaca MCF variables for model.csv') %>% 
  mutate(mortality_qual= fct_relevel(mortality_qual, levels=c("None", "Low", "Moderate", "High"))) %>% 
  # mutate(dnbr_qualitative = case_when(
  #   dnbr < 0.1 ~ "unburned",
  #   dnbr < 0.27 ~ "low",
  #   dnbr < 0.66 ~ "moderate",
  #   dnbr > 0.66 ~ "high")) %>% 
  # mutate(dnbr_qualitative = as.factor(dnbr_qualitative),
  #        dnbr_qualitative = fct_relevel(dnbr_qualitative, levels=c("high", "moderate", "low", "unburned"))) %>% 
  mutate(aspect_transformed = (1-cos((2*pi*aspect)/360))/2) %>% 
  mutate(three_class_mortality = case_when(
    mortality_qual %in% "High" ~ "High",
    mortality_qual %in% "Moderate" ~ "Moderate",
    mortality_qual %in% c("Low", "None") ~ "Low")) %>% 
  mutate(three_class_mortality = as.factor(three_class_mortality)) %>% 
  mutate(three_class_mortality= fct_relevel(three_class_mortality, 
                                     levels=c("Low", "Moderate", "High")))

@@

******WOULD LIKE TO GET DNBR AND RDNBR TO COMPARE TO OUR ESTIMATES FOR MORTALITY!!*******

@@

@@

@@

@@

#random forest code---------------------------------------------------
for (i in 1:50)
{

@@

  #random forest model
  rfm <- randomForest(tc_mortality ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + ridge + drainage_association + ndvi, data = train, ntree=1500, importance=T, nscale=0)

@@

  # lists to save model outputs
  test_accuracy_list <- append(test_accuracy_list, test_accuracy)
  write.table(model_importance, file="rfm model tc importance.csv", sep=",",col.names=NA, append=T) 
  write.table(tabb, file = "mcf tc test accuracy table.csv", sep=",",col.names=NA, append=T)
}
